---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# mpoly

__mpoly__ is a simple collection of tools to help deal with multivariate polynomials _symbolically_ and functionally in R.  Polynomials are defined with the `mp` function:

```{r mp, echo = TRUE}
library(mpoly)
mp("x + y")
mp("(x + 4y)^2 (x - .25)")
```

[Term orders](http://en.wikipedia.org/wiki/Lexicographical_order#Monomials) are available with the reorder function:
```{r}
(p <- mp("(x + y)^2 (1 + x)"))
reorder(p, varorder = c('y','x'), order = 'lex')
reorder(p, varorder = c('x','y'), order = 'glex')
```

Arithmetic is defined for them (`+`, `-`, `*` and `^`):
```{r}
p1 <- mp("x + y")
p2 <- mp("x - y")

p1 + p2
p1 - p2
p1 * p2
p1^2
```


You can compute derivatives easily:
```{r}
deriv(mp("x + x y + x y^2"), "y")
```

And turn them into functions you can evaluate:
```{r}
f <- as.function(mp("(x-2) x (x+2)"))
s <- seq(-2.5, 2.5, .1)
df <- expand.grid(x = s)
df$f <- apply(df, 1, f)

library(ggplot2)
qplot(x, f, data = df, geom = "line")
```



```{r}
f <- as.function(mp("x^2 - y^2")) 
df <- expand.grid(x = s, y = s)
df$f <- apply(df, 1, f)
qplot(x, y, data = df, geom = "tile", fill = f)
```

I'm starting to put in methods for some other R functions:

```{r}
n <- 101
df <- data.frame(x = seq(-5, 5, length.out = n))
df$y <- with(df, -x^2 + 2*x - 3 + rnorm(n, 0, 2))

mod <- lm(y ~ x + I(x^2), data = df)
(p <- round(as.mpoly(mod)))
qplot(x, y, data = df) +
  stat_function(fun = as.function(p), colour = 'red')

```

Grobner bases computations are available using __rSymPy__:

```{r}
polys <- mp(c("t^4 - x", "t^3 - y", "t^2 - z"))
grobner(polys)
```


## Installation

* From CRAN: `install.packages("mpoly")`

* From Github (dev version): 
```R
# install.packages("devtools")
devtools::install_github("Rexamine/stringi")
devtools::install_github("hadley/stringr")
devtools::install_github("dkahle/mpoly")
```



